Overlapping Batches for the Assessment of Solution Quality in Stochastic Programs

### 2014-07-15 omrp_lb_rev.tex
figures --gb
Crossing the t's, dotting the i's

\begin{abstract}
\noindent Overlapping Batch Means (OBM) has long been used in simulation as a method of reusing data to generate variance estimators with asymptotically lower variance.
In this paper, we apply the OBM method to stochastic programming by formulating a variant of the multiple replications procedure used for assessing solution quality.
We give conditions under which the resulting optimality gap estimators are strongly consistent and have the same asymptotically lower variance as the classical OBM estimators \citep{Meketon1984,Welch1987}.
Computational experiments for several test problems are presented, examining the small-sample behavior and the computational efficiency of the overlapping batches method in this context.\medskip

@@

A natural setting where batch means are used in stochastic optimization (especially in stochastic programming) is for assessing solution quality. 
This typically involves generating point and interval estimators of the optimality gap of a given candidate solution using Monte Carlo simulation. 
Because optimality gap estimators can be difficult to analyze statistically, a common approach is to use nonoverlapping batches of data to generate a number of optimality gap estimators. 
These individual estimators are then used to form point and interval estimators applying the nonoverlapping batch means method.
This is referred to as the Multiple Replications Procedure (MRP) and was initially discussed in \citep{Mak1999}.   
In this paper, we formulate an overlapping batches version of MRP with a parametrized degree of overlap.
%, and present theoretical and computational results.

@@

\item[(iv)] As the degree of overlap increases, the number of batches and optimization problems solved increases because we need to solve a sampling problem per batch. 
At the same time, variances of the estimators are lowered, and when warm starting is available, solution time per batch can decrease, resulting in efficiency gains. 
We explore this trade-off between computation time and lower variances.
% and efficiency gains. 
%We explore the overall computation time versus lower variances and efficiency gains trade-off. 
Our results indicate that when warm-starting is effective, overlapping is computationally more efficient relative to estimators formed via nonoverlapping batches.
%Our results indicate that overlapping can indeed increase computational efficiency, especially when warm-starting is effective. 
We provide guidelines on selecting the degree of overlap based on computational efficiency and demonstrate efficiency on a newsvendor problem. 
\end{itemize}

@@

Overlapping batch means reuses a subset of the observations from one batch to form the next batch. 
As an example, suppose each batch contains $m$ observations and that there are a total of $n=2m$ observations.
In the nonoverlapping case, there are only two batches:
% the first batch contains observations $y^1, y^2, \dots, y^m$ and the second batch contains observations $y^{m+1}, y^{m+2}, \dots, y^{2m}$.  
$y^1, y^2, \dots, y^m$ and $y^{m+1}, y^{m+2}, \dots, y^{2m}$.  
In the overlapping batches of \citet{Meketon1984}, the first batch is as before, the second batch now consists of $y^2, y^3, \dots, y^{m+1}$, the third batch contains observations $y^3, y^4, \dots, y^{m+2}$, and so on.  
This is what we call the \emph{maximally-overlapping} batch means method, because $m-1$ observations are common to---and only one observation changes between---adjacent batches.  
It is possible to vary the amount of overlap between neighboring batches. 
This generalization was first studied by \citet{Welch1987}, who also established the relationship between OBM and spectral estimators; see also \citep{Song1992}.
%
The overlapping batch means 
%formed using the batch nonoverlap parameter $1 \leq \gamma < m$ 
are no longer independent. 
However, by simply reusing the data in this fashion one obtains a better variance estimator. 
That is, the overlapping variance estimator has (i) nearly the same bias as the standard nonoverlapping variance estimator, and it has (ii)
%the overlapping estimator has 
a lower asymptotic variance than its nonoverlapping counterpart.

In this paper, we use the {\it batch nonoverlap parameter} $1 \leq \gamma \leq m$ to denote how much neighboring batches do {\it not} overlap.  
For instance, $\gamma = m$ corresponds to the case of nonoverlapping batches and $\gamma = 1$ corresponds to the maximally-overlapping case of \citet{Meketon1984}.  
In general, $\gamma$ can take any integer value between and including $[1,m]$. 
Let $\lfloor \cdot \rfloor$ be the greatest integer that is less than or equal to its argument. 
The sample mean of each batch is calculated similarly, 
$
\yb_j = \frac{1}{m} \sum_{i=1}^m y^{\gamma(j-1) + i},\ \  j = 1, 2, \dots, \nb, 
$
now taking into account $\gamma$.  
Here, $\nb$ batches are formed given $n$, $m$, and $\gamma$.  
The variance estimator is updated to
\begin{equation} \label{eq:variance_gamma}
	\vartg{\ybb} = \frac{1}{\left( \tfrac{n}{m} - 1 \right) \left( \nb \right)}  \sum_{j=1}^{\nb} (\yb_j - \ybb)^2.
\end{equation}
With $\gamma=m$ and $n=mk$, (\ref{eq:variance_gamma}) reduces to (\ref{eq:var}).
With $\gamma=1$ (the maximally-overlapping case), the denominator in (\ref{eq:variance_gamma}) is slightly different than the one in  \citep{Meketon1984}.
This slightly different denominator 
makes $\vto{\ybb}$ an unbiased estimator for i.i.d.\ data for all finite $m$ and $n$ with the same asymptotic benefits \citep{Song1992}.

The amount of asymptotic reduction in $\var{\vartg{\ybb}}$ for a given $0\leq \gamma< m$ relative its nonoverlapping counterpart 
%$\var{\vtm{\ybb}}$ 
depends on the asymptotic ratio of $\gamma/m$, which we denote by $\gammab$. 
The maximally-overlapping case ($\gamma = 1$) corresponds to $\gammab=0$, and the nonoverlapping case corresponds to $\gammab=1$. 
\cite{damerdji1994strong,damerdji1995mean} studied batching structures that guarantee almost sure and mean-square consistency of $\vartg{\ybb}$ in the nonoverlapping and maximally-overlapping cases. 
Under appropriate conditions, the results of \cite{damerdji1995mean}, along with those of \cite{Meketon1984} and \cite{Welch1987}, establish that 
\begin{equation}
\label{eq:varvarred}
\frac{n}{m}\var{n \vartg{\ybb}} \rightarrow \beta_{\gammab} \sigma^4
\end{equation}
as the batch size $m$ and then the number of batches $n/m$ tend to infinity. 
In (\ref{eq:varvarred}), $\beta_1 = 2$ for nonoverlapping batches, $\beta_0 = \frac{4}{3}$ for maximally overlapping batches, and more generally, $\beta_{\gammab}$ is twice the value of 
\begin{equation} \label{eq:var_reduct_formula}
	\gammab \left( 1 + 2 \sum_{j=1}^{\cogb-1} (1-j\gammab)^2 \right),
\end{equation}
where $\lceil \cdot \rceil$ denotes the smallest integer that is greater than or equal to its argument.
If $\ogb$ is an integer, (\ref{eq:var_reduct_formula}) simplifies to $\frac{2+\gammab^2}{3}$.
This means that the maximally-overlapping variance estimator has a variance that is two-thirds ($66.67\%$) of the nonoverlapping estimator.
As the amount of overlap decreases, the variance increases. 
When only half of the observations overlap ($\gammab = 1/2$ or $\gamma=m/2$), for instance, the variance is 75\% of the nonoverlapping counterpart.

@@

Another difference between our setting and simulation output analysis is that once the data are generated through a simulation, they can be reused without much additional computational effort to obtain the OBM variance estimator.  
In this setting, the computational effort may increase with data reuse because we need to solve a sampling problem (SP$_m$) for each batch.  
The increase in computation time can be alleviated in two ways: (i) by increasing $\gamma$ to partially overlap the batches, resulting in a smaller number of batches, and (ii) using warm starting to quickly solve the sampling approximations with overlapping samples. 
%We use the batch nonoverlap parameter $\gamma$ as before to allow for varying degrees of overlap in OMRP. 
We examine the effect of warm starting and any resulting computational efficiency gains in \S \ref{ssec:compeff}. 
We are now ready to define the OMRP estimators.

%First, near-optimal variance reduction can be obtained by partially overlapping the batches \citep{Welch1987,Song1992}.  
%Partial overlapping results in a smaller number of batches; hence, fewer optimization problems need to be solved.
%Second, in many solution methods, warm-starting can be used to quickly solve the sampling approximations with overlapping samples, considerably reducing solution time.
%In this paper, we consider a parametrized degree of overlap, where the degree of overlap is at the user's discretion, and examine this both theoretically and empirically. 
% THIS LAST SENTENCE NEEDS TO CHANGE!

@@

In this section, we provide conditions under which several asymptotic properties hold. 
First, we show the consistency of the point estimators of optimality gap, $\gbb_l$, and variance, $n_l VG_l$. 
Then, we provide conditions under which the same benefits on variance reduction are observed in the stochastic programming setting by overlapping. 
Finally, we show the asymptotic validity of the OMRP interval estimator.

@@

First, we restrict our attention to i.i.d.\ sampling by A1.
The results for overlapping batches from the simulation output analysis indicate that $\var{n_l VG_l}$ may be reduced by overlapping, thus requiring at least finite fourth moments. 
We will formally show this result in Section~\ref{ssec:var_reduct}.
Our proof relies on an exponential rate of convergence of the sampled solution to the true solution of the problem, specified by A3.  
Assumption A3 can be satisfied for different classes of problems with unique optimal solutions.  
\citet{shapiro2000rate} establish exponential rates of convergence for a class of (\ref{eq:sto_prog}), which include two-stage stochastic linear programs with recourse when the support of $\xit$ is finite.   
\citet{kleywegt2002sample} show that assumption A3 can be satisfied for discrete stochastic programs (i.e., when $X$ is a finite set) under the condition that the moment generating function of $f(\xs,\xit) - f(\x,\xit)$ for $\x \in X \backslash \{\xs\}$ is finite valued on $\R$.  
This class of problems includes stochastic integer programs with recourse.
  We present computational results on both stochastic linear and integer programs with recourse in \S \ref{sec:comp}.

@@

@@

We are now ready to show that the resulting point estimators are strongly consistent. 
We use the abbreviation a.s.\ to denote almost sure convergence. 
We also use $\mu_{\xh} := \e{f(\xh,\xit) - f(\xs,\xit)}$ to denote the optimality gap of $\xh$ and $\sigma^2_{\xh} := \var{f(\xh,\xit) - f(\xs,\xit)}$ to denote the associated variance term.

@@

Theorem \ref{thm:varvar_conv} gives conditions under which the same asymptotically lower variances are achieved for the OMRP optimality gap variance estimator relative to its nonoverlapping counterpart MRP. 
We will empirically examine the variance of OMRP estimator for small sample sizes in \S \ref{sec:comp}. 
A hypothesis of Theorem \ref{thm:varvar_conv} is a batching structure that ensures convergence of $\frac{n_l}{m_l}\var{n_l VD_l}$ to the appropriate $\beta_{\gammab} \sigma^4_{\xh}$, ensuring the mean-square consistency of $n_l VD_l$. 
\citet{damerdji1995mean} shows that mean-square convergence is guaranteed by $\tfrac{1}{2} + \tfrac{1}{4+\epsilon} < r < 1$ for both nonoverlapping and maximally overlapping batches under the batching structure $m_l = \lfloor n_l^r \rfloor$ and $4+\epsilon$ finite moments assumption A2$^\prime$.  
With A2, the range of $r$ that ensures mean-square consistency can be found by setting $\epsilon=0$.

@@

The overlapping variance estimator for stochastic programs displays the same desirable properties as in simulation, but we must still check that it results in a valid CI.  
Let $I^G_l = \gbb_l + t_{d_l,\alpha} \sqrt{VG_l}$ be the width of the one-sided CI generated by OMRP. 
The degrees of freedom $d_l$ is the inverse of (\ref{eq:var_reduct_formula}) using $\frac{\gamma_l}{m_l}$ instead of $\gammab$.

@@

A summary of results from these experiments is depicted in Figures~\ref{fig:varvar1}--\ref{fig:nv}.
Figures~\ref{fig:varvar1} and \ref{fig:varvar2} show the reduction of variance, with each term being normalized with respect to $\var{nVG}$ of the nonoverlapping case (MRP). 
The solid lines in these figures show the theoretical variance reduction from \citep{Welch1987} given by (\ref{eq:var_reduct_formula}). 
Empirical results agree with the theoretical values well, with more variability observed in CEP1, 10D, and DB1. 
These experiments suggest that similar variance reduction can be achieved in stochastic programming by overlapping the batches, even with small sample sizes.
The variance reduction in DB1 can be less than the asymptotic amount %for the small sample sizes used in these experiments. 
because DB1 has not yet been solved to optimality due its size; therefore, it may lack a unique optimal solution. 
%, violating assumption A3. (?)
Even with the relatively small sample sizes compared to the size of this problem, however, overlapping appears to achieve some variance reduction.

@@

Figure~\ref{fig:nv}(a) shows estimates of $\e{nVG}$ in the newsvendor problem for changing batch size and amount of overlap. 
We can see that the expectation of the variance estimator is not changed with increasing overlap. 
Although the estimate of $\e{nVG}$ looks large for $m=100$, the error from $\sigma^2_\xh$ is never more than 1\%.  
Estimates for $\e{nVG}$ showed a similar pattern for CEP1, 10D, and DB1.  
Those graphs are omitted for brevity.

Finally, Figure~\ref{fig:nv}(b) shows the coverage probability of CIs generated by OMRP for several values of $m$ across varying values of $\gamma/m$ for the newsvendor problem.  
The results of the overlapping batches estimators from the simulation literature show that coverage probability does not change with the amount of overlap, which we can  empirically see in this figure for OMRP for NVD.
We observed the same for CEP1, 10D, and DB1.  
The coverage probabilities from applying the (nonoverlapping) MRP algorithm to these problems presented in \citep{Bayraksan2006} agree with our results: for the newsvendor problem, coverage probability drops as $m$ increases and the coverage probability of CEP1 (not shown for brevity) remains fairly constant around the desired value of $90\%$. 
Coverage probability for 10D (not shown) and {\it estimated} coverage probability for DB1 (also not shown) were very high, more than $0.99$.  
The bias from solving (\ref{eq:sto_prog_m}) for these problems seems to be much more significant than the variance.

@@

The increased computational workload from solving overlapping batches raises the question of the computational efficiency of this method to estimate $VG$.
In this setting, computational efficiency can be defined as $[T(\gammab)\var{VG}]^{-1}$, where $T(\gammab)$ is the computational time spent to generate $VG$ for a given value of $\gammab$.
Because the variance of the overlapping variance estimator relative to the nonoverlapping (MRP) variant is known from (\ref{eq:var_reduct_formula}), we need only examine the computational time, which can be decomposed as $T(\gammab) = t(\gammab)\left(\nb\right)$. 
Here, $t(\gammab)$ is the average computation time per batch, which is multiplied by the number of batches.
Warm-starting an algorithm may be used to decrease the computation time per batch for $\gammab < 1$, leading to a possible increase in computational efficiency.

@@

\begin{figure}[thb!]
	\centering
	\subfloat[][Theoretical]{
	   \label{fig:efficiency_theory}
	   \includegraphics[width=.5\textwidth]{images/comp_efficiency_theory}}
	\subfloat[][Empirical for NVD]{
	   \label{fig:efficiency_nv}
	   \includegraphics[width=.5\textwidth]{images/nv_comp_efficiency}}
		\caption{
		Computational efficiency of the OMRP variance estimator relative to the (nonoverlaping) MRP variance estimator measured
		(a) theoretically and
		(b) empirically for NVD using insertion sort
%		(graphs are scaled so that nonoverlapping batches produces a computational efficiency of 1)
		}
\label{fig:efficiency}
\end{figure}

@@

Another related question on computational efficiency would be whether OMRP is more efficient than nonoverlapping MRP with a larger number of batches.
When generating observations is cheap, one could alternatively increase the number of batches in MRP to obtain estimators with lower variances. 
In the nonoverlapping case, $\var{VG} \propto \frac{1}{k}$; so, decreasing $\var{VG}$ to the level of maximally overlapping batches of \cite{Meketon1984} requires a $50\%$ increase in the number of batches, and thus a $50\%$ increase in computation time.
Continuing on in this way for various values of $\gammab$, we obtain the solid black curve in Figure \ref{fig:comp_work_comparison}. 
This curve shows the increase in computational effort due to an increased number of batches in MRP in order to obtain the same variance for OMRP across $\gammab$.
If the work increase by OMRP is less than this solid black curve, then OMRP is computationally more efficient than increasing the number of batches in MRP. 
Again, assuming average computation time per batch can be represented as $T(\gammab) \propto \gammab^p$, Figure \ref{fig:comp_work_comparison}(a) compares the increase in computational work for OMRP for several values of $p$ to the work increase for MRP to get an equivalent decrease in variance.
We again see that OMRP is advantageous for larger values of $p$.
For small values of $p$, some efficiency gain is observed at $\gammab$ values close to 1.
We again tested this empirically on NVD using insertion sort with $m=1,000$, average of $10,000$ independent runs. 
The results, shown in Figure \ref{fig:comp_work_comparison}(b), again appear to match the theory well.

\begin{figure}[htb!]
	\centering
	\subfloat[][Theoretical]{
	   \label{fig:comp_work_comparison_theory}
	   \includegraphics[width=.5\textwidth]{images/work_comparison}}
	\subfloat[][Empirical for NVD]{
	   \label{fig:comp_work_comparison_nv}
	   \includegraphics[width=.5\textwidth]{images/nv_work_comparison}}
		\caption{
		Comparison of computational work required to get equivalent variance between MRP variance estimator with increased number of batches and OMRP variance estimator measured
		(a) theoretically and 
		(b) empirically for NVD using insertion sort
%		(Left) A theoretical comparison of the computational work required to get equivalent variance reduction between OMRP and MRP when the number of batches are increased.
%		(Right) Measured computational work for NVD compared with the theoretical results.
		}
\label{fig:comp_work_comparison}
\end{figure}

>>>>

\begin{abstract}
\noindent Overlapping Batch Means (OBM) has long been used in simulation as a method of reusing data to generate variance estimators with asymptotically lower variance.
In this paper, we apply the OBM method to stochastic programming by formulating a variant of the multiple replications procedure used for assessing solution quality.
We give conditions under which the resulting optimality gap point estimators are strongly consistent, interval estimators are asymptotically valid, and the variance estimators have the same asymptotically lower variances relative to their nonoverlapping counterparts as the classical OBM estimators \citep{Meketon1984,Welch1987}.
Computational experiments for several test problems are presented, examining the small-sample behavior and the computational efficiency of the overlapping batches method in this context.\medskip

@@

A natural setting where batch means are used in stochastic optimization (especially in stochastic programming) is for assessing solution quality. 
This typically involves generating point and interval estimators of the optimality gap of a given candidate solution using Monte Carlo simulation. 
Because optimality gap estimators can be difficult to analyze statistically, a common approach is to use nonoverlapping batches of data to generate a number of optimality gap estimators. 
These individual estimators are then used to form point and interval estimators applying the nonoverlapping batch means method.
This is referred to as the Multiple Replications Procedure (MRP) and was initially discussed in \citep{Mak1999}.   
In this paper, we formulate an overlapping batches version of MRP.

@@

\item[(iv)] As the degree of overlap increases, the number of batches and optimization problems solved increases because we need to solve a sampling problem per batch. 
At the same time, variances of the estimators are lowered, and when warmstarting is available, solution time per batch can decrease, resulting in efficiency gains. 
We explore this trade-off between computation time and lower variances.
% and efficiency gains. 
%We explore the overall computation time versus lower variances and efficiency gains trade-off. 
Our results indicate that when warmstarting is effective, overlapping is computationally more efficient relative to estimators formed via nonoverlapping batches.
%Our results indicate that overlapping can indeed increase computational efficiency, especially when warmstarting is effective. 
We provide guidelines on selecting the degree of overlap based on computational efficiency and demonstrate efficiency on a newsvendor problem. 
\end{itemize}

@@

Overlapping batch means reuses a subset of the observations from one batch to form the next batch. 
As an example, suppose each batch contains $m$ observations and that there are a total of $n=2m$ observations.
In the nonoverlapping case, there are only two batches:
% the first batch contains observations $y^1, y^2, \dots, y^m$ and the second batch contains observations $y^{m+1}, y^{m+2}, \dots, y^{2m}$.  
$y^1, y^2, \dots, y^m$ and $y^{m+1}, y^{m+2}, \dots, y^{2m}$.  
In the overlapping batches of \citet{Meketon1984}, the first batch is as before, the second batch now consists of $y^2, y^3, \dots, y^{m+1}$, the third batch $y^3, y^4, \dots, y^{m+2}$, and so on.  
This is what we call the \emph{maximally-overlapping} batch means method, because $m-1$ observations are common to---and only one observation changes between---adjacent batches.  
It is possible to vary the degree of overlap between neighboring batches. 
This generalization was first studied by \citet{Welch1987}, who also established the relationship between OBM and spectral estimators; see also \citep{Song1992}.
%
The overlapping batch means 
%formed using the batch nonoverlap parameter $1 \leq \gamma < m$ 
are no longer independent. 
However, by simply reusing the data in this fashion one obtains a better variance estimator. 
That is, the overlapping variance estimator has (i) nearly the same bias as the standard nonoverlapping variance estimator, and it has (ii)
%the overlapping estimator has 
a lower asymptotic variance than its nonoverlapping counterpart.

In this paper, we use the {\it batch nonoverlap parameter} $1 \leq \gamma \leq m$ to denote how much neighboring batches do {\it not} overlap.  
For instance, $\gamma = m$ corresponds to the case of nonoverlapping batches and $\gamma = 1$ corresponds to the maximally-overlapping case of \citet{Meketon1984}.  
In general, $\gamma$ can take any integer value between and including $[1,m]$. 
Let $\lfloor \cdot \rfloor$ be the greatest integer that is less than or equal to its argument. 
The sample mean of each batch is calculated similarly, 
$
\yb_j = \frac{1}{m} \sum_{i=1}^m y^{\gamma(j-1) + i},\ \  j = 1, 2, \dots, \nb, 
$
now taking into account $\gamma$.  
Here, $\nb$ batches are formed given $n$, $m$, and $\gamma$.  
The variance estimator is updated to
\begin{equation} \label{eq:variance_gamma}
	\vartg{\ybb} = \frac{1}{\left( \tfrac{n}{m} - 1 \right) \left( \nb \right)}  \sum_{j=1}^{\nb} (\yb_j - \ybb)^2.
\end{equation}
With $\gamma=m$ and $n=mk$, (\ref{eq:variance_gamma}) reduces to (\ref{eq:var}).
With $\gamma=1$ (the maximally-overlapping case), the denominator in (\ref{eq:variance_gamma}) is slightly different than the one in  \citep{Meketon1984}.
This slightly different denominator 
makes $\vto{\ybb}$ an unbiased estimator for i.i.d.\ data for all finite $m$ and $n$ with the same asymptotic benefits of nearly the same bias and lower variances relative to its counterpart obtained via nonoverlapping batches \citep{Song1992}.

The amount of asymptotic reduction in $\var{\vartg{\ybb}}$ for a given $0\leq \gamma< m$ relative its nonoverlapping counterpart 
%$\var{\vtm{\ybb}}$ 
depends on the asymptotic ratio of $\gamma/m$, which we denote by $\gammab$. 
The maximally-overlapping case ($\gamma = 1$) corresponds to $\gammab=0$, and the nonoverlapping case corresponds to $\gammab=1$. 
\cite{damerdji1994strong,damerdji1995mean} studied batching structures that guarantee almost sure and mean-square consistency of $\vartg{\ybb}$ in the nonoverlapping and maximally-overlapping cases. 
Under appropriate conditions, the results of \cite{damerdji1995mean}, along with those of \cite{Meketon1984} and \cite{Welch1987}, establish that 
\begin{equation}
\label{eq:varvarred}
\frac{n}{m}\var{n \vartg{\ybb}} \rightarrow \beta_{\gammab} \sigma^4
\end{equation}
as the batch size $m$ and then the number of batches $n/m$ tend to infinity. 
In (\ref{eq:varvarred}), $\beta_1 = 2$ for nonoverlapping batches, $\beta_0 = \frac{4}{3}$ for maximally overlapping batches, and more generally, $\beta_{\gammab}$ is twice the value of 
\begin{equation} \label{eq:var_reduct_formula}
	\gammab \left( 1 + 2 \sum_{j=1}^{\cogb-1} (1-j\gammab)^2 \right),
\end{equation}
where $\lceil \cdot \rceil$ denotes the smallest integer that is greater than or equal to its argument.
If $\ogb$ is an integer, (\ref{eq:var_reduct_formula}) simplifies to $\frac{2+\gammab^2}{3}$.
This means that the maximally-overlapping variance estimator has a variance that is two-thirds ($66.67\%$) of the nonoverlapping estimator.
As the degree of overlap decreases, the variance increases. 
When only half of the observations overlap ($\gammab = 1/2$ or $\gamma=m/2$), for instance, the variance is 75\% of the nonoverlapping counterpart.

@@

Another difference between our setting and simulation output analysis is that once the data are generated through a simulation, they can be reused without much additional computational effort to obtain the OBM variance estimator.  
In this setting, the computational effort may increase with data reuse because we need to solve a sampling problem (SP$_m$) for each batch.  
The increase in computation time can be alleviated in two ways: (i) by increasing $\gamma$ to partially overlap the batches, resulting in a smaller number of batches, and (ii) using warmstarting to quickly solve the sampling approximations with overlapping samples. 
%We use the batch nonoverlap parameter $\gamma$ as before to allow for varying degrees of overlap in OMRP. 
We examine the effect of warmstarting and any resulting computational efficiency gains in \S \ref{ssec:compeff}. 
We are now ready to define the OMRP estimators.

%First, near-optimal variance reduction can be obtained by partially overlapping the batches \citep{Welch1987,Song1992}.  
%Partial overlapping results in a smaller number of batches; hence, fewer optimization problems need to be solved.
%Second, in many solution methods, warmstarting can be used to quickly solve the sampling approximations with overlapping samples, considerably reducing solution time.
%In this paper, we consider a parametrized degree of overlap, where the degree of overlap is at the user's discretion, and examine this both theoretically and empirically. 
% THIS LAST SENTENCE NEEDS TO CHANGE!

@@

In this section, we provide conditions under which several asymptotic properties hold for OMRP estimators. 
First, we show the consistency of the point estimators of optimality gap, $\gbb_l$, and variance, $n_l VG_l$. 
Then, we provide conditions under which the asymptotically lower variances are observed for the OMRP variance estimator relative to nonoverlapping MRP estimator.  
Finally, we show the asymptotic validity of the OMRP interval estimator.

@@

First, we restrict our attention to i.i.d.\ sampling by A1.
The results for overlapping batches from the simulation output analysis indicate that $\var{n_l VG_l}$ may be reduced by overlapping, thus requiring at least finite fourth moments. 
We will formally show this result in \S~\ref{ssec:var_reduct}.
Our proof relies on an exponential rate of convergence of the sampled solution to the true solution of the problem, specified by A3.  
Assumption A3 can be satisfied for different classes of problems with unique optimal solutions.  
\citet{shapiro2000rate} establish exponential rates of convergence for a class of (\ref{eq:sto_prog}), which include two-stage stochastic linear programs with recourse when the support of $\xit$ is finite.   
\citet{kleywegt2002sample} show that assumption A3 can be satisfied for discrete stochastic programs (i.e., when $X$ is a finite set) under the condition that the moment generating function of $f(\xs,\xit) - f(\x,\xit)$ for $\x \in X \backslash \{\xs\}$ is finite valued on $\R$.  
This class of problems includes stochastic integer programs with recourse.
  We present computational results on both stochastic linear and integer programs with recourse in \S \ref{sec:comp}.

@@

Before we begin establishing the theoretical results, we introduce the following notation. 
We use $\mu_{\xh} := \e{f(\xh,\xit) - f(\xs,\xit)}$ to denote the optimality gap of $\xh$ and $\sigma^2_{\xh} := \var{f(\xh,\xit) - f(\xs,\xit)}$ to denote the associated variance term.
With i.i.d.\ sampling, $\e{\db_{lj}}=\e{\dbb_l}=\mu_{\xh}, \ \forall l,j$ and by (\ref{eq:varvarred}), $\frac{n_l}{m_l} \var{n_l VD_l} \rightarrow \beta_{\gammab} \sigma^4_{\xh}$ as $l \rightarrow \infty$ under appropriate conditions.

@@

We are now ready to show that the resulting point estimators are strongly consistent. 
We use the abbreviation a.s.\ to denote almost sure convergence.

@@

Theorem \ref{thm:varvar_conv} gives conditions under which the same asymptotically lower variances are achieved for the OMRP optimality gap variance estimator relative to its nonoverlapping counterpart, the MRP variance estimator. 
We will empirically examine the variance of OMRP estimator for small sample sizes in \S \ref{sec:comp}. 
A hypothesis of Theorem \ref{thm:varvar_conv} is a batching structure that ensures convergence of $\frac{n_l}{m_l}\var{n_l VD_l}$ to the appropriate $\beta_{\gammab} \sigma^4_{\xh}$, ensuring the mean-square consistency of $n_l VD_l$. 
\citet{damerdji1995mean} shows that mean-square convergence is guaranteed by $\tfrac{1}{2} + \tfrac{1}{4+\epsilon} < r < 1$ for both nonoverlapping and maximally overlapping batches under the batching structure $m_l = \lfloor n_l^r \rfloor$ and $4+\epsilon$ finite moments assumption A2$^\prime$.  
With A2, the range of $r$ that ensures mean-square consistency can be found by setting $\epsilon=0$.

@@

The overlapping variance estimator for stochastic programs displays the same desirable properties, detailed in \S \ref{subsec:conv} and \S \ref{ssec:var_reduct}, as in simulation, but we must still check that it results in a valid CI.  
Let $I^G_l = \gbb_l + t_{d_l,\alpha} \sqrt{VG_l}$ be the width of the one-sided CI generated by OMRP. 
The degrees of freedom $d_l$ is the inverse of (\ref{eq:var_reduct_formula}) using $\frac{\gamma_l}{m_l}$ instead of $\gammab$.

@@

A summary of results from these experiments is depicted in Figures~\ref{fig:varvar1}--\ref{fig:nv}.
Figures~\ref{fig:varvar1} and \ref{fig:varvar2} show the variance of OMRP variance estimators, normalized with respect to $\var{nVG}$ of the nonoverlapping case (MRP). 
The solid lines in these figures show the theoretical relative variance from \citep{Welch1987} given by (\ref{eq:var_reduct_formula}). 
Empirical results agree with the theoretical values well, with more variability observed in CEP1, 10D, and DB1. 
These experiments suggest that lower variances can be achieved in stochastic programming by overlapping the batches, even with small sample sizes.
The variance of OMRP variance estimators in DB1 can be less than the asymptotic amount established by Theorem \ref{thm:varvar_conv}. %for the small sample sizes used in these experiments. 
This is because DB1 has not yet been solved to optimality due its size; therefore, it may lack a unique optimal solution. 
%, violating assumption A3. (?)
Even with the relatively small sample sizes compared to the size of this problem, however, overlapping appears to reduce the variances.

@@

Figure~\ref{fig:nv}(a) shows estimates of $\e{nVG}$ in the newsvendor problem for changing batch size and degree of overlap. 
We can see that the expectation of the variance estimator is not changed with increasing overlap. 
Although the estimate of $\e{nVG}$ looks large for $m=100$, the error from $\sigma^2_\xh$ is never more than 1\%.  
Estimates for $\e{nVG}$ showed a similar pattern for CEP1, 10D, and DB1.  
Those graphs are omitted for brevity.

Finally, Figure~\ref{fig:nv}(b) shows the coverage probability of CIs generated by OMRP for several values of $m$ across varying values of $\gamma/m$ for the newsvendor problem.  
The results of the overlapping batches estimators from the simulation literature show that coverage probability does not change with the degree of overlap, which we can  empirically see in this figure for OMRP for NVD.
We observed the same for CEP1, 10D, and DB1.  
The coverage probabilities from applying the (nonoverlapping) MRP algorithm to these problems presented in \citep{Bayraksan2006} agree with our results: for the newsvendor problem, coverage probability drops as $m$ increases and the coverage probability of CEP1 (not shown for brevity) remains fairly constant around the desired value of $90\%$. 
Coverage probability for 10D (not shown) and {\it estimated} coverage probability for DB1 (also not shown) were very high, more than $0.99$.  
The bias from solving (\ref{eq:sto_prog_m}) for these problems seems to be much more significant than the variance.

@@

The increased computational workload from solving overlapping batches raises the question of the computational efficiency of this method to estimate $VG$.
In this setting, computational efficiency can be defined as $[T(\gammab)\var{VG}]^{-1}$, where $T(\gammab)$ is the computational time spent to generate $VG$ for a given value of $\gammab$.
Because the variance of the overlapping variance estimator relative to the nonoverlapping (MRP) variant is known from (\ref{eq:var_reduct_formula}), we need only examine the computational time, which can be decomposed as $T(\gammab) = t(\gammab)\left(\nb\right)$. 
Here, $t(\gammab)$ is the average computation time per batch, which is multiplied by the number of batches.
Warmstarting an algorithm may be used to decrease the computation time per batch for $\gammab < 1$, leading to a possible increase in computational efficiency.

@@

\begin{figure}[thb!]
	\centering
	\subfloat[][Theoretical]{
	   \label{fig:efficiency_theory}
	   \includegraphics[width=.49\textwidth]{images/comp_efficiency_theory}}
	\subfloat[][Empirical for NVD]{
	   \label{fig:efficiency_nv}
	   \includegraphics[width=.51\textwidth]{images/nv_comp_efficiency}}
		\caption{
		Computational efficiency of the OMRP variance estimator relative to the (nonoverlaping) MRP variance estimator measured
		(a) theoretically and
		(b) empirically for NVD using insertion sort
%		(graphs are scaled so that nonoverlapping batches produces a computational efficiency of 1)
		}
\label{fig:efficiency}
\end{figure}

@@

A related question on computational efficiency asks whether OMRP is more efficient than nonoverlapping MRP with a larger number of batches.
When generating observations is cheap, one could alternatively increase the number of batches in MRP to obtain estimators with lower variances. 
In the nonoverlapping case, $\var{VG} \propto \frac{1}{k}$; so, decreasing $\var{VG}$ to the level of maximally overlapping batches of \cite{Meketon1984} requires a $50\%$ increase in the number of batches, and thus a $50\%$ increase in computation time.
Continuing on in this way for various values of $\gammab$, we obtain the solid black curve in Figure \ref{fig:comp_work_comparison}. 
This curve shows the increased computational effort in MRP due to an increased number of batches to obtain the same variance of OMRP across $\gammab$.
If the work increase by OMRP is less than this solid black curve, then OMRP is computationally more efficient than increasing the number of batches in MRP. 
Again, assuming average computation time per batch can be represented as $T(\gammab) \propto \gammab^p$, Figure \ref{fig:comp_work_comparison}(a) compares the increase in computational work for OMRP for several values of $p$ to the work increase for MRP to get an equivalent decrease in variance.
We again see that OMRP is advantageous for larger values of $p$.
For small values of $p$, some efficiency gain is observed at $\gammab$ values close to 1.
We again tested this empirically on NVD using insertion sort with $m=1,000$, taking an average of $10,000$ independent runs. 
The results, shown in Figure \ref{fig:comp_work_comparison}(b), appear to match the theory well.

\begin{figure}[htb!]
	\centering
	\subfloat[][Theoretical]{
	   \label{fig:comp_work_comparison_theory}
	   \includegraphics[width=.51\textwidth]{images/work_comparison}}
	\subfloat[][Empirical for NVD]{
	   \label{fig:comp_work_comparison_nv}
	   \includegraphics[width=.49\textwidth]{images/nv_work_comparison}}
		\caption{
		Comparison of computational work required to get equivalent variance between MRP variance estimator with increased number of batches and OMRP variance estimator measured
		(a) theoretically and 
		(b) empirically for NVD using insertion sort
%		(Left) A theoretical comparison of the computational work required to get equivalent variance reduction between OMRP and MRP when the number of batches are increased.
%		(Right) Measured computational work for NVD compared with the theoretical results.
		}
\label{fig:comp_work_comparison}
\end{figure}
